Modeling the Linux page cache for accurate simulation of data-intensive applications

### 2020-12-14 paper_io_sim.tex
Edits in results, started editing conclusion

            We also implemented our model as part of \wrench, enhancing its
            internal implementation and APIs with a page cache abstraction,
            and allowing users to activate the feature via a command-line
            argument. For the experiments in this paper we use \wrench 1.6
            at commit
            \href{https://github.com/wrench-project/wrench/tree/67185374330d2c4bf274fce222c937e838df5b03}{6718537433},
            which uses \simgrid 3.25, available at
            \url{https://framagit.org/simgrid/simgrid}. Our implementation
            is now part of \wrench's master branch and will be available to
            users with the upcoming \wrench 1.8 release. \wrench provides a full \simgrid-based simulation 
            environment that supports, among other features, concurrent accesses to storage devices, 
            applications distributed on multiple hosts, network transfers, 
            and multi-threading.

@@

        The synthetic application, implemented in C, consists of three single-core,
        sequential tasks where each task reads the file produced by the
        previous task, increments every byte of this file to emulate real
        processing, and writes the resulting data to disk. In real
        executions, we clear the page cache before starting the application,
        to ensure comparable conditions among the runs.
        The anonymous memory used by application
        is released after each task, which we also simulate in the Python prototype and in WRENCH-cache.

@@

        The page cache simulation model drastically reduces I/O simulation
        errors in the three successive application tasks (Figure~\ref{fig:single_error}). The first read is not impacted
        as it only involves uncached data. Errors are reduced from an average
        of 345\% in the original \wrench to 46\% in the Python prototype and
        39\% in \wrench-cache. Unsurprisingly, the original \wrench simulator
        significantly overestimates read and write times, due to its lack
        of page cache simulation model. Results with files of 50~GB and 75~GBs
        show similar behaviors and are not reported due to space limitations.

@@

        Simulated memory profiles are highly consistent with the real ones
        (Figure~\ref{fig:single_memprof}). With 20~GB files, memory
        profiles almost exactly match the real ones, although dirty data
        seems to be flushing faster in real life than in simulation
        \tristan{any idea why?}, a behavior also observed with 100~GB files. With
        100~GB files, used memory reaches total memory during the first write,
        triggering dirty data flushing. Used memory drops to cached memory
        when application tasks release anonymous memory. Simulated cached
        memory is highly consistent with real values, except toward the end
        of Read 3 where it slightly increases in simulation but not in
        reality. Our explanation could be that in reality, because File 3 is
        entirely in cache, the cache amount does not increase when File 3 is read.
        While in simulation, a part of File 3 is read from disk, leading to an increase
        in cache amount followed by a decrease when eviction is called.
        Apart from that, dirty data remains under the dirty ratio as expected. 
        The Python prototype and \wrench-cache exhibit nearly
        identical memory profiles, which reinforces the confidence in our
        implementation.

        The content of the simulated memory cache is also highly consistent
        with reality (Figure~\ref{fig:single_cache}). With 20~GB files, the
        simulated cache content exactly matches reality, since all files fit in page cache. With 100~GB files, a slight
        discrepancy is observed after Write 2, which explains the
        simulation error previously mentioned in Read 3. In the real execution indeed,
        File 3 is entirely cached after Write 2, whereas in the simulated execution, 
        only a part of it is cached. 
        This is due to the fact that the in simulation, data is flushed and evicted 
        from the inactive list first, then from the active list. Thus, File 2, which is 
        accessed again by Read 2 and is cached in the active list, is not flushed and evicted, 
        while File 3 is flushed and evicted since it is accessed once only and cached 
        in the inactive list. 
        This flushing and eviction order in simulation may not correctly reflect the 
        order in reality and lead to the difference.

@@

            \begin{figure*}
            \centering
            \begin{subfigure}{\linewidth}
                \centering
                   \includegraphics[width=\linewidth]{result/single/figures/single_errors.pdf}
                   \vspace*{-0.7cm}
                   \caption{Absolute relative simulation errors}
                   \vspace*{0.5cm}
                   \label{fig:single_error}
                \end{subfigure}
            \begin{subfigure}{\linewidth}
                \centering
                %    Gray shades represent task phases (read, compute and write).
                %    Lines represent memory usage along pipeline execution time.}
                   \includegraphics[width=\linewidth]{result/single/figures/single_memprof.pdf}
                   \vspace*{-0.7cm}
                   \caption{Memory profiles \tristan{Remove trailing whitespace for 20GB graphs (xmax should be about 160s, not 200s)}}
                   \vspace*{0.5cm}
                   \label{fig:single_memprof}
            \end{subfigure}
            \begin{subfigure}{\linewidth}
                \centering
                   \includegraphics[width=1.05\linewidth]{result/single/figures/cached_files.pdf}
                   \caption{Cache contents \emph{after} application I/O operations}
                %    \textcolor{red}{Update real results of 20GB}}
                   \label{fig:single_cache}
            \end{subfigure}
            \caption{Single-threaded results}
            \end{figure*}

        \subsection{Concurrent applications}

@@

            The page cache model notably reduces \wrench's simulation error
            for concurrent applications and local I/Os
            (Figure~\ref{fig:multi_local}). For reads, \wrench-cache still
            slightly overestimates reality, due to the discrepancy between
            simulated and real read bandwidths mentioned before. For writes,
            \wrench-cache slightly underestimates reality for the same reason. The plateau
            ending around 7 concurrent pipelines in the \wrench-cache writes
            marks the limit beyond which the page cache is saturated with dirty data
            and needs flushing \tristan{why is there no plateau in real life?}.

        \subsection{Remote storage}

@@

        \subsection{Overhead comparison}
        As is the case for \wrench, simulation time with \wrench-cache scales
        linearly with the number of concurrent applications
        (Figure~\ref{fig:multi_time}, p \textless $10^{-24}$). However, the page
        cache model substantially increases the simulation time by
        application as can be seen by comparing regression slopes in
        Figure~\ref{fig:multi_time}. Interestingly, \wrench-cache is faster with 
        NFS I/Os than with local I/Os, most likely due to the use of writethrough
        cache in NFS, which by-passes flushing operations.
        \begin{figure}
            \begin{subfigure}{\columnwidth}
                \centering
                \includegraphics[width=\linewidth]{result/multi/figures/simulation_time.pdf}
            \end{subfigure}
            \caption{Simulation time comparison. \wrench-cache scales
            linearly with the number of concurrent applications, albeit
            with a higher overhead than \wrench.}
            \label{fig:multi_time}
            \end{figure}

    \section{Conclusion}
    \label{discussion}
        Page cache undoubtedly has important impacts on performance, in
        particular for data-intensive applications. Numerical simulation is
        an effective approach to predict and compare the performance of
        applications, schedulers and infrastructures. However, page cache
        is missing in many simulation tools for distributed systems. In
        this study, we proposed a simulation model of I/Os with page cache,
        we implemented it in the \simgrid-based \wrench library, and we
        evaluated its benefits in different conditions. Results show that
        our model very substantially improves simulation accuracy,
        decreasing absolute relative errors between simulated and real
        performance by a factor close to 10 (see results of the
        single-threaded experiment). 
        %Furthermore, this level of accuracy
        %can be achieved without a fine-grained level of details
        %\tristan{not sure what you mean}. 
        % Dzung: I meant the results are accurate without a very detailed model. But the overhead results dot not seem to show that.
        Our page cache model is available
        in \wrench from commit \href{https://github.com/wrench-project/wrench/tree/d4e188d2f0cf69a19ae0a6b5e41530eb77d12cd8}{d4e188d2f0}, ready to be used 
        for the simulation of Big Data applications.

        Simulation results could be made even more accurate by a deeper
        investigation of page cache flushing and eviction order. 
        The availability of asymmetrical bandwidths in the
        forthcoming \simgrid release will also directly improve our
        simulation accuracy. Relevant future work includes simulating
        application memory consumption, random I/O, file readahead and
        persistent disk storage.

>>>>

            We also implemented our model as part of \wrench, enhancing its
            internal implementation and APIs with a page cache abstraction,
            and allowing users to activate the feature via a command-line
            argument. We used SimGrid's locking mechanism to handle
            concurrent accesses to page cache LRU lists by the two Memory
            Manager threads. For the experiments in this paper we use
            \wrench 1.6 at commit
            \href{https://github.com/wrench-project/wrench/tree/67185374330d2c4bf274fce222c937e838df5b03}{6718537433},
            which uses \simgrid 3.25, available at
            \url{https://framagit.org/simgrid/simgrid}. Our implementation
            is now part of \wrench's master branch and will be available to
            users with the upcoming \wrench 1.8 release. \wrench provides a full \simgrid-based simulation 
            environment that supports, among other features, concurrent accesses to storage devices, 
            applications distributed on multiple hosts, network transfers, 
            and multi-threading.

@@

        The synthetic application, implemented in C, consists of three single-core,
        sequential tasks where each task reads the file produced by the
        previous task, increments every byte of this file to emulate real
        processing, and writes the resulting data to disk. Files are
        numbered by ascending creation times: File 1 is file read by Task 1, and File \texttt{i+1} is the file
        written by task \texttt{i}. In real executions, we clear the page
        cache before starting the application, to ensure comparable
        conditions among the runs. The anonymous memory used by application
        is released after each task, which we also simulate in the Python
        prototype and in WRENCH-cache.

@@

        The page cache simulation model drastically reduces I/O simulation
        errors in each application task (Figure~\ref{fig:single_error}). The first read is not impacted
        as it only involves uncached data. Errors are reduced from an average
        of 345\% in the original \wrench to 46\% in the Python prototype and
        39\% in \wrench-cache. Unsurprisingly, the original \wrench simulator
        significantly overestimates read and write times, due to the lack
        of page cache simulation. Results with files of 50~GB and 75~GB
        show similar behaviors and are not reported for brevity.

@@

        Simulated memory profiles are highly consistent with the real ones
        (Figure~\ref{fig:single_memprof}). Used memory consists of both
        cached and anonymous memory that cannot be evicted at the current
        time. With 20~GB files, memory profiles almost exactly match the
        real ones, although dirty data seems to be flushing faster in real
        life than in simulation \tristan{any idea why?}, a behavior also
        observed with 100~GB files. With 100~GB files, used memory reaches
        total memory during the first write, triggering dirty data
        flushing. Used memory drops to cached memory when application tasks
        release anonymous memory. Simulated cached memory is highly
        consistent with real values, except toward the end of Read 3 where
        it slightly increases in simulation but not in reality. This
        occurred due to the fact that after Write 2, File 3 was only partly
        cached in simulation whereas it was entirely cached in the real
        system. In all cases, dirty data remains under the dirty ratio as
        expected. The Python prototype and \wrench-cache exhibit nearly
        identical memory profiles, which reinforces the confidence in our
        implementation.

        The content of the simulated memory cache is also highly consistent
        with reality (Figure~\ref{fig:single_cache}). With 20~GB files, the
        simulated cache content exactly matches reality, since all files fit in page cache. With 100~GB files, a slight
        discrepancy is observed after Write 2, which explains the
        simulation error previously mentioned in Read 3. In the real execution indeed,
        File 3 is entirely cached after Write 2, whereas in the simulated execution, 
        only a part of it is cached. 
        This is due to the fact that the in simulation, data is flushed and evicted 
        from the inactive list first, then from the active list \tristan{whereas in the real execution ...?}. Thus, File 2, which is 
        accessed again by Read 2 and is cached in the active list, is not flushed and evicted, 
        while File 3 is flushed and evicted since it is accessed once only and cached 
        in the inactive list.

@@

            \begin{figure*}
            \centering
            \begin{subfigure}{\linewidth}
                \centering
                   \includegraphics[width=\linewidth]{result/single/figures/single_errors.pdf}
                   \vspace*{-0.7cm}
                   \caption{Absolute relative simulation errors}
                   \vspace*{0.5cm}
                   \label{fig:single_error}
                \end{subfigure}
            \begin{subfigure}{\linewidth}
                \centering
                %    Gray shades represent task phases (read, compute and write).
                %    Lines represent memory usage along pipeline execution time.}
                   \includegraphics[width=\linewidth]{result/single/figures/single_memprof.pdf}
                   \vspace*{-0.7cm}
                   \caption{Memory profiles \tristan{Remove trailing
                   whitespace for 20GB graphs (xmax should be about 160s,
                   not 200s)} \tristan{Find a way to add Read 1, Write 1,
                   Read 2, etc on the graphs, perhaps at the top of the
                   real execution graphs}}
                   \vspace*{0.5cm}
                   \label{fig:single_memprof}
            \end{subfigure}
            \begin{subfigure}{\linewidth}
                \centering
                   \includegraphics[width=1.05\linewidth]{result/single/figures/cached_files.pdf}
                   \caption{Cache contents \emph{after} application I/O operations}
                %    \textcolor{red}{Update real results of 20GB}}
                   \label{fig:single_cache}
            \end{subfigure}
            \caption{Single-threaded results}
            \end{figure*}

        \subsection{Concurrent application (local storage)}

@@

            The page cache model notably reduces \wrench's simulation error
            for concurrent applications executed with local I/Os
            (Figure~\ref{fig:multi_local}). For reads, \wrench-cache still
            slightly overestimates reality, due to the discrepancy between
            simulated and real read bandwidths mentioned before. For writes,
            \wrench-cache slightly underestimates reality for the same reason. The plateau
            ending around 7 concurrent applications in the \wrench-cache writes
            marks the limit beyond which the page cache is saturated with dirty data
            and needs flushing \tristan{why is there no plateau in real life?}.

        \subsection{Remote storage (NFS)}

@@

        \subsection{Overhead comparison}
        As is the case for \wrench, simulation time with \wrench-cache scales
        linearly with the number of concurrent applications
        (Figure~\ref{fig:multi_time}, p \textless $10^{-24}$). However, the page
        cache model substantially increases simulation time by
        application, as can be seen by comparing regression slopes in
        Figure~\ref{fig:multi_time}. Interestingly, \wrench-cache is faster with 
        NFS I/Os than with local I/Os, most likely due to the use of writethrough
        cache in NFS, which bypasses flushing operations.
        \begin{figure}
            \begin{subfigure}{\columnwidth}
                \centering
                \includegraphics[width=\linewidth]{result/multi/figures/simulation_time.pdf}
            \end{subfigure}
            \caption{Simulation time comparison. \wrench-cache scales
            linearly with the number of concurrent applications, albeit
            with a higher overhead than \wrench.}
            \label{fig:multi_time}
            \end{figure}

    \section{Conclusion}
    \label{discussion}
    We designed a model of the Linux page cache and implemented it in
    the \simgrid-based \wrench library to simulate distributed
    applications. Evaluation results show that our model very
    substantially improves simulation accuracy, decreasing absolute
    relative simulation errors up to 9$\times$ (see results of
    the single-threaded experiment).
    Our page cache model is publicly available
    in the \wrench GitHub repository.

    Page cache simulation would be instrumental in a number of studies.
    For instance, it is now common for HPC clusters to run 
    application jobs in Linux control groups (cgroups), where resource consumption 
    including memory is controlled.

    With the page cache, I/O times are strongly dependent on the amount of
    available memory, which is itself impacted by the use of anonymous
    memory by applications. A more accurate simulation of application
    memory consumption would therefore be a relevant future direction to
    further improve I/O simulation.

    I/Os remain much more complex than is currently available in simulation
    models.

        Simulation results could be made even more accurate by a deeper
        investigation of page cache flushing and eviction order. 
        The availability of asymmetrical bandwidths in the
        forthcoming \simgrid release will also directly improve our
        simulation accuracy.

        Relevant future work includes simulating
        application memory consumption, random I/O, file readahead and
        persistent disk storage.
